Using VariantSpark to learn population structure from genomic profiles

### 2015-10-26 doc/publications/Recomb15/recomb15.tex
more touchups

\subsubsection{Random Forest does not scale to whole genome analysis and is outperformed by logistic regression}
In this section we compare RF against \kMeans\ and the LR.  
\marginpar{TODO elaborate on the error Aidan: See near end of paragraph} 
Despite using {\sc Spark} and {\sc MLlib}, RF is not able to scale to the full dataset of 1000 individuals and 40
Million variables, due to a limitation of the Java virtual machine (maximum array size) that causes {\sc Mllib}'s
random forest implementation to fail. This error is due to the number of features present in each sample,
with a potential solution being to limit the features through feature selection, or simply using a subset of the dataset.
We therefore use only chromosome 22 (494,328 alleles) for the comparison in this section.  We train a RF
and LR on this subset of genomic profiles from the 1000 Genomes data using the annotated labels of all
super-populations.  We perform 10-fold cross-validation and report the mean error over all training and testing folds
along with the standard error of the mean (SEM). We also use the Adjusted Rand Index as a metric for model quality
as it allows us to compare the classification results with the \kMeans{} clustering model. We also calculate the error
for the two classification methods, which is simply the number of misclassified individuals divided by the number of samples.
We perform 10 runs of \kMeans{} with 4 clusters (\(k=4\)) and report the mean ARI of the
resulting models along with the standard error.

\marginpar{TODO elaborate on accuracy measures Aidan: See above}
As reported in our earlier work, \kMeans\ is not able to fit the dataset accurately.
Although capable of fitting a relatively accurate model (\(ARI=0.837\)), the average ARI over ten models is  \(0.596 \pm 0.067\), see table~\ref{RF}.
RF achieves a better fitting model with a mean ARI of \(0.769 \pm 0.021\) but is outperformed by the simple logistic regression, with an ARI of \(0.923 \pm 0.015\).
This indicates that a multivariate approach may not be needed to accurately fit ethnicity.

@@

The mean ARI for our \kMeans clustering models is \(0.960 \pm 0.004 \). Although slightly lower, this value is similar to the deep-learning approach
where they achieved an ARI of 0.976 (where one ASW and one GBR individual is misclassified as CHB). Our supervised logistic regression approach,
however, improves upon this result.  With this approach, we achieve an accuracy of 100\%. This is the average across
ten folds, with a standard error of 0.

Although we see two of our ten runs meet the accuracy of the deep-learning approach, where they achieved an accuracy of
99\% (or ARI of 0.975, with two misclassified individuals), our logistic regression approach demonstrates a consistent
improvement.  With this approach, we achieve an accuracy of \(99.997 \pm 0.003\) across ten folds.

@@

\section{Conclusion}
In this paper we systematically investigated the performance of supervised and unsupervised methods for determining the
ethnicity from whole genome variant data.  We found that a supervised method, logistic regression, outperforms the
unsupervised approach, \kMeans{} (XX vs YY).  We also found that ethnicity can be better predicted by the simple
logistic regression method compared to the multivariate approaches such as random forest or deep learning as reported by
Big Data Genomics using the ADAM framework (XX vs Y1 and Y2, respectively).  Furthermore, we observe a limitation within
Java, which limits the random forest implementation of {\sc Spark}'s {\sc MLlib} to XX features for the 1069 individuals
of the 1000 genome project.  This property would limit the 100,000 genome project to using only YY features for their
70,000 individuals, when using the same random forest implementation.  We conclude that while logistic regression
outperforms all other methods it remains unable to accurately classify individuals from [Sub-populations] from the AMR
super-population.  We therefore ague that the remaining mixture in X is reflective of the migrational background and
hence the underlying genomic profile of Y is accurately similar to X.  This may be an important observation for medical
applications that need accurate population association.

>>>>

\subsubsection{Random Forest does not scale to whole genome analysis and is outperformed by logistic regression}
In this section we compare RF against \kMeans\ and the LR.  
\marginpar{TODO elaborate on the error Aidan: See near end of paragraph} 
Despite using {\sc Spark} and {\sc MLlib}, RF is not able to scale to the full dataset of 1000 individuals and 40
Million variables, due to a limitation of the Java virtual machine (maximum array size) that causes {\sc Mllib}'s
random forest implementation to fail. This error is due to the number of features present in each sample,
with a potential solution being to limit the features through feature selection, or simply using a subset of the dataset.
We therefore use only chromosome 22 (494,328 alleles) for the comparison in this section.  We train a RF
and LR on this subset of genomic profiles from the 1000 Genomes data using the annotated labels of all
super-populations.  We perform 10-fold cross-validation and report the mean error over all training and testing folds
along with the standard error of the mean (SEM). We also use the Adjusted Rand Index (ARI) as a metric for model quality
as it allows us to compare the classification results with \kMeans{} clustering models. We also calculate the error
for the two classification methods, which is simply the number of misclassified individuals divided by the number of samples.
We perform 10 runs of \kMeans{} with 4 clusters (\(k=4\)) and report the mean ARI of the
resulting models along with the standard error.

\marginpar{TODO elaborate on accuracy measures Aidan: See above paragraph}
As reported in our earlier work, \kMeans\ is not able to fit the dataset accurately.
Although capable of fitting a relatively accurate model (\(ARI=0.837\)), the average ARI over ten models is  \(0.596 \pm 0.067\), see table~\ref{RF}.
RF achieves a better fitting model with a mean ARI of \(0.769 \pm 0.021\), however it's outperformed by the simple logistic regression, with an ARI of \(0.923 \pm 0.015\).
This indicates that a multivariate approach may not be needed to accurately fit ethnicity.

@@

The mean ARI for our \kMeans clustering models is \(0.960 \pm 0.004 \). Although slightly lower, this value is similar to the deep-learning approach
where they achieved an ARI of 0.976 (where one ASW and one GBR individual is misclassified as CHB). Our supervised logistic regression approach,
however, improves upon this result.  With this approach, we achieve an accuracy of 100\%. This is the average across
ten folds.

@@

\section{Conclusion}
In this paper we systematically investigated the performance of supervised and unsupervised methods for determining the
ethnicity from whole genome variant data.  We found that a supervised method, logistic regression, outperforms the
unsupervised approach, \kMeans{} (0.936 vs 0.84).  We also found that ethnicity can be better predicted by the simple
logistic regression method compared to the unsupervised \kMeans{} approach or the multivariate deep learning approach as reported by
Big Data Genomics using the ADAM framework (1 vs 0.96 and 0.976, respectively).  Furthermore, we observe a limitation within
Java, which limits the random forest implementation of {\sc Spark}'s {\sc MLlib} to XX features for the 1069 individuals
of the 1000 genome project.  This property would make feature selection mandatory for the 100,000 genome project when classifying their
70,000 individuals, when using the same random forest implementation.  We conclude that while logistic regression
outperforms all other methods it remains unable to accurately classify individuals from [Sub-populations] from the AMR
super-population.  We therefore ague that the remaining mixture in X is reflective of the migrational background and
hence the underlying genomic profile of Y is accurately similar to X.  This may be an important observation for medical
applications that need accurate population association.
